{\bf {\Large Improving estimates of ecosystem metabolism computed from dissolved oxygen time series}}

### 2014-11-04 abstract.tex
some edits from Jim's comments, windows specific to each case study

In aquatic ecosystems, time series of \ac{DO} can be used to infer integrated ecosystem processes such as primary production, respiration, and net metabolism.  However, \ac{DO} time series data at estuaries may reflect variation from both biological and physical processes, potentially leading to inaccurate or misleading ecosystem process estimates.  One such physical process is the occurrence of large lateral \ac{DO} gradients in an estuary which may advect water with different \ac{DO} characteristics past a sensor.  In such situations, the lateral gradient may cause variation in \ac{DO} time series that are not attributable to metabolic processes.  Statistical techniques that dynamically quantify variation in \ac{DO} and tidal changes over time have the potential to isolate biological signals in \ac{DO} variation.  A weighted regression method was developed to filter the \ac{DO} time series to remove the influence of physical advection, thereby removing bias or noise in ecosystem metabolism estimates.  The method was tested using simulated \ac{DO} time series with known additive components of biological and physical variation.  The method was validated using one year of continuous monitoring data at four water quality stations that are part of the \acl{NERRS}. We provide a detailed discussion on use of the method for improving certainty in ecosystem metabolism estimates from sites with strong tidal influences.  This approach could improve metabolism estimates using shorter deployment periods or incomplete time series by removing bias attributed to lateral water movement.

>>>>

In aquatic ecosystems, time series of \ac{DO} have been used to compute estimates of integrated ecosystem metabolism.  Central to this open water or ``Odum'' method is the assumption that the dissolved oxygen time series is a Lagrangian specification of the flow field.  However, most \ac{DO} time series are collected at fixed locations, such the method must assume that changes in dissolved oxygen principally reflect ecosystem metabolism and that effects due to advection or mixing can be neglected.  A statistical model using weighted regression was applied to separate variability in \ac{DO} associated with metabolism from tidal variation or other advection in estuaries, thereby helping to partially relax this assumption and improve estimates of ecosystem metabolism. The method was developed and tested using a simulated \ac{DO} time series with known biological and physical components, and then applied to one year of continuous monitoring data from four water quality stations within the \acl*{NERRS}.  Overall, the approach is a useful way to reduce variability in estimates of ecosystem metabolism caused by advection, particularly when the magnitude of tidal influence is high. By reducing the effects of physical transport on metabolism estimates, there may be increased potential to empirically relate metabolic rates to causal factors on times scales of several days to several weeks. Estimates of variability associated with physical advection may also be more interpretable, since convolution of physical and biological effects can be reduced, providing increased potential to resolve differences in metabolism across a mosaic of estuarine habitats.

### 2014-11-04 assess.Rnw
some edits from Jim's comments, windows specific to each case study

A point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  The effects of tidal variation on daily estimates may not be relevant if seasonal or annual aggregations remove this potential bias.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after filtering were generally unchanged for the two-week period. \cref{tab:case_res} also indicated that mean annual estimates of production and respiration were unchanged for Rookery Bay and Sapelo Island.  However, annual averages of production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay. Given these results, tidal variation may or may not have effects on metabolism estimates on aggregated time scales longer than 24 hours, depending on the location.  Therefore, an evaluation of weighted regression to filter the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for its application.  Specifically, when should filtering be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and filtered estimates that are aggregated over different periods of observation (e.g., annual, seasonal, monthly) could help address this question.

The observed and filtered daily estimates were averaged by month and season (Fall, Spring, Summer, and Winter) for each case study to evaluate effects of aggregation on mean production and respiration.  Mean annual estimates in \cref{tab:case_res} also provided a basis of comparison with monthly and seasonal aggregation. Significant variation in aggregated production and respiration estimates for month and season was observed for each case study (\cref{fig:metab_sum1,fig:metab_sum2}).  Filtered production and respiration estimates for Padilla Bay and Rookery Bay exhibited seasonal and monthly variation that was more characteristic of expected trends during warmer months.  Specifically, production estimates based on observed \ac{DO} were substantially muted for both Padilla Bay (\cref{fig:metab_sum1}) and Rookery Bay (\cref{fig:metab_sum2}) during summer months, whereas values were significantly higher after filtering. Results for Sapelo Island suggested that winter and summer months were under- and over-estimated, respectively, based on the observed data.  Results for Elkhorn Slough varied significantly such that production and respiration were significantly reduced after filtering regardless of the aggregation period.  Overall, these trends emphasize the importance of considering different aggregation periods for interpreting metabolism estimates.  Each case study showed differences in observed and filtered values at monthly and seasonal aggregations, whereas only two of the four case studies had mean aggregated estimates that were substantially different (Elkhorn Slough and Padilla Bay, \cref{tab:case_res}).  Periods of observation as long as one year may include significant sources of bias from tidal advection, suggesting the need for applying weighted regression given careful consideration of appropriate window widths.

>>>>

A point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  The effects of tidal variation on daily estimates may not be relevant if seasonal or annual aggregations (e.g., mean annual metabolism) remove this potential bias.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after filtering were generally unchanged for the two-week period. Alternatively, annual averages of production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay (\cref{tab:case_res}). Given these results, tidal variation may or may not have effects on metabolism estimates on aggregated time scales, depending on the location.  Therefore, an evaluation of weighted regression to filter the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for its application.  Specifically, when should filtering be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and filtered estimates that are aggregated over different periods of observation could help address this question.

The observed and filtered daily estimates were averaged by month for each case study to evaluate effects of aggregation on mean production and respiration.  Mean annual estimates in \cref{tab:case_res} also provided a basis of comparison with the aggregations. Significant variation in aggregated production and respiration estimates was observed for each case study (\cref{fig:metab_sum}).  Filtered production and respiration estimates for Padilla Bay and Rookery Bay exhibited monthly variation that was more characteristic of expected trends during warmer months.  Specifically, metabolism estimates based on observed \ac{DO} were substantially muted for both Padilla Bay and Rookery Bay during summer months, whereas values were significantly higher after filtering. Results for Sapelo Island suggested that aggregated estimates were similar before and after filtering, although winter and summer months were slightly under- and over-estimated, respectively, using the observed data.  Results for Elkhorn Slough varied significantly such that production and respiration were significantly reduced after filtering.  Overall, these trends emphasize the importance of considering different aggregation periods for interpreting metabolism estimates.  Each case study showed differences in observed and filtered values at monthly aggregations, whereas only two of the four case studies had mean aggregated estimates that were substantially different (Elkhorn Slough and Padilla Bay, \cref{tab:case_res}).  Periods of observation as long as one year may include significant sources of bias from tidal advection, suggesting the need for applying weighted regression given careful consideration of appropriate window widths.

### 2014-11-04 intro.Rnw
some edits from Jim's comments, windows specific to each case study

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  The fundamental assumption is that the time series of \ac{DO} describes the same water mass over time \citep{Needoba12}.  Estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey13}.  Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Application of the method to estuaries is a particular concern as physical mixing caused by tidal currents may confound the biological variation in \ac{DO} time series \citep{Kemp80,Caffrey03,Nidzieko14}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from phyical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting significant variation from advection of water masses with different metabolic histories was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Murrell13}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} acknowledged this need by exploring the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the application of a method for filtering an observed \ac{DO} time series for estimated tidal effects to more accurately quantify estimates of ecosystem metabolism for estuaries.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  Specifically, one year of \ac{DO} time series for each case study was filtered to adjust estimates of ecosystem metabolism to apparent tidal effects.  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate measure that approximates lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.  Overall, the analysis is meant to better characterize the relative roles of biological and physical processes in estuarine systems.

>>>>

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection.  
A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey13}.  Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Application of the method to estuaries is a particular concern as physical mixing caused by tidal currents may confound the biological variation in \ac{DO} time series \citep{Kemp80,Caffrey03,Nidzieko14}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from physical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Murrell13}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} are caused by metabolism associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate characterization of lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.
